---
title: "Getting Started with jsfR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with jsfR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # Python jsf may not be available on all build machines
)
```

## Overview

The **jsfR** package provides an R interface to the Python
[`jsf`](https://github.com/DGermano8/jsf) package using
[`reticulate`](https://rstudio.github.io/reticulate/). It implements the
**Jump-Switch-Flow (JSF)** algorithm — a hybrid stochastic-deterministic
simulator for compartmental models.

The core idea is simple: when a compartment's population is **large**, it is
simulated via a fast deterministic ODE (mean-field). When it is **small**, it
switches to an exact Continuous-Time Markov Chain (CTMC / Gillespie SSA). The
switching happens automatically at a user-defined `SwitchingThreshold`,
giving you the speed of ODEs and the accuracy of stochastic simulation exactly
where it matters — near zero.

```
Population > Threshold  →  ODE (fast, deterministic)
Population ≤ Threshold  →  CTMC (exact, stochastic)
```

This makes JSF particularly powerful in **infectious disease modelling**, where
the number of infected individuals can drop near zero, and stochastic extinction
events (i.e., disease elimination) are scientifically critical.

---

## Installation

```{r install, eval = FALSE}
# Install the package
devtools::install_github("Vai-Man/jsf-tests", subdir = "jsfR")

# Install the Python jsf dependency
jsfR::install_jsf()
```

---

## The `JSFSimResult` S7 Object

All simulations in `jsfR` are run through `jsf_run()`, which returns a typed
`JSFSimResult` S7 object. This object:

- Stores compartment trajectories under **named** list entries
- Records simulation metadata (`method`, `dt`, `SwitchingThreshold`, `seed`)
- Provides a `print()` method for a tidy diagnostic summary
- Provides a `plot()` method for a `ggplot2` time-series

---

## Example 1: Birth-Death Process

The simplest compartmental model, a single population with birth rate
$\lambda$ and death rate $\mu$.

$$\frac{dN}{dt} = \lambda N - \mu N \qquad \implies \qquad E[N(t)] = N_0\, e^{(\lambda - \mu)t}$$

Where:

- $N$: Population size
- $\lambda$: Birth rate
- $\mu$: Death rate

```{r birth-death}
library(jsfR)

rates_bd <- function(x, t) c(1.0 * x[[1]],   # birth
                              0.5 * x[[1]])    # death

stoich_bd <- list(
  nu         = list(c(1L), c(-1L)),
  DoDisc     = c(1L),
  nuReactant = list(c(1L), c(1L)),
  nuProduct  = list(c(2L), c(0L))
)

result_bd <- jsf_run(
  x0                  = 5L,
  rates               = rates_bd,
  stoich              = stoich_bd,
  t_max               = 6.0,
  switching_threshold = 10L,
  enforce_do          = 0L,
  seed                = 42L,
  compartment_names   = "Population"
)
```

The `print()` method shows a per-compartment diagnostic summary:

```{r print-bd}
print(result_bd)
#> JSFSimResult
#>   Method    : operator-splitting
#>   dt        : 0.01
#>   Threshold : 10
#>   Seed      : 42
#>   Time range: [0.000, 6.000]  (547 steps)
#>   Population             init=   5.0  final= 142.0  min=   3.0  max= 142.0
```

The `plot()` method returns a `ggplot2` object:

```{r plot-bd}
plot(result_bd)
```

---

## Example 2: SIS Infectious Disease Model

The SIS (Susceptible-Infected-Susceptible) model is the canonical use case for
JSF, directly cited in the official package documentation. It models a disease
that can eliminate stochastically when the infected count drops near zero —
something deterministic ODEs cannot capture.

The mathematical formulation for this model is:
$$\frac{dS}{dt} = -\beta SI + \gamma I$$
$$\frac{dI}{dt} = \beta SI - \gamma I$$

Where:

- $S$: Susceptible population size
- $I$: Infected population size
- $\beta$: Transmission rate
- $\gamma$: Recovery rate

We track `Infected` discretely (`DoDisc = c(0L, 1L)`) because extinction is
only possible in the stochastic regime.

```{r sis}
rates_sis <- function(x, t) c(
  0.002 * x[[1]] * x[[2]],   # S + I -> 2I  (infection)
  1.0   * x[[2]]              # I -> S       (recovery)
)

stoich_sis <- list(
  nu         = list(c(-1L,  1L), c( 1L, -1L)),
  DoDisc     = c(0L, 1L),        # only Infected tracked stochastically
  nuReactant = list(c(1L, 1L),   c(0L, 1L)),
  nuProduct  = list(c(0L, 2L),   c(1L, 0L))
)

result_sis <- jsf_run(
  x0                  = c(997L, 3L),
  rates               = rates_sis,
  stoich              = stoich_sis,
  t_max               = 10.0,
  switching_threshold = c(1000L, 15L),
  enforce_do          = c(0L, 0L),
  seed                = 1L,
  compartment_names   = c("Susceptible", "Infected")
)

print(result_sis)
plot(result_sis)
```

The `DoDisc` vector is the key control: `c(0L, 1L)` forces `Susceptible` into
the fast ODE regime (it stays large) while keeping `Infected` stochastic at
low counts. This is what allows the simulation to capture disease extinction
accurately without sacrificing performance on the large `Susceptible` pool.

---

## Comparing SSA vs JSF

`jsfR` also ships with a native R implementation of the **Gillespie SSA**
(`simulate_birth_death()`) so you can compare both algorithms side-by-side.
This is useful for validating that the hybrid JSF method matches the exact
stochastic simulation in expectation.

```{r compare}
library(ggplot2)

# Run 100 Gillespie SSA replicates
ssa_raw  <- simulate_birth_death_ensemble(lambda = 1.0, mu = 0.5, x0 = 5L,
                                          t_max = 6.0, n_reps = 100L, seed = 42L)

# Run 100 JSF replicates
jsf_raw  <- generate_jsf_reference(lambda = 1.0, mu = 0.5, x0 = 5L,
                                    t_max = 6.0, n_reps = 100L, seed = 42L)

# Interpolate to a common time grid and compute metrics
ssa_grid <- interpolate_ensemble(ssa_raw)
jsf_grid <- interpolate_ensemble(jsf_raw)
comp     <- compute_comparison(ssa_grid, jsf_grid)

print_comparison_summary(comp)
#> SSA vs JSF Comparison Summary:
#>   Time range         : [0.00, 6.00]
#>   Mean absolute error: 3.1842
#>   Max  absolute error: 38.6210
#>   Mean relative error: 0.0793
#>   RMSE               : 5.6091
```

Both methods agree closely in expectation, confirming that JSF is a valid
approximation of the exact stochastic process while running significantly
faster at high population levels.

---

## Further Reading

- [JSF Python documentation](https://github.com/DGermano8/jsf)
- [Jump-Switch-Flow paper (Germano et al.)](https://github.com/DGermano8/jsf)
- `?jsf_run`: full API reference for the main simulation function
- `?simulate_birth_death`: Gillespie SSA implementation
- `?JSFSimResult`: S7 class reference
