---
title: "Jump-Switch-Flow (JSF) Demonstration"
subtitle: "Using the Python jsf package via reticulate in R"
author: "Vaibhav Manihar"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    self-contained: true
---

## Overview

The **Jump-Switch-Flow (JSF)** algorithm is a hybrid stochastic-deterministic 
simulator for compartmental models. It couples Continuous Time Markov Chains 
(CTMCs) with Ordinary Differential Equations (ODEs), allowing compartments to 
dynamically switch between discrete ("jumping") and continuous ("flowing") 
representations based on population size thresholds.

This document demonstrates how to use the Python `jsf` package from R via 
`reticulate`.

## Setup

We first configure `reticulate` to use an appropriate Python environment, 
install `jsf` if needed, and import required packages.

```{r setup, message=FALSE, warning=FALSE}
# Load R packages
library(reticulate)
library(ggplot2)

# Determine if we have a local copy of jsf, otherwise install from GitHub
local_jsf <- file.path(getwd(), "jsf")

if (dir.exists(local_jsf)) {
  jsf_pkg <- normalizePath(local_jsf, winslash = "/")
  # Ensure the local module is prioritised in Python's path to avoid shadowing
  sys <- import("sys")
  py_run_string(sprintf("import sys\nsys.path.insert(0, r'%s')", jsf_pkg))
} else {
  jsf_pkg <- "git+https://github.com/DGermano8/jsf.git"
}

# Install required numerical libraries into the active Reticulate environment
py_install(c("numpy", "pandas", "matplotlib", "python-libsbml"), pip = TRUE)
if (!py_module_available("jsf")) {
  py_install(jsf_pkg, pip = TRUE)
}

# Clear any cached imports to ensure the correct module is loaded
py_run_string("import sys\nif 'jsf' in sys.modules:\n    del sys.modules['jsf']")

# Import Python modules into R
jsf    <- import("jsf")
random <- import("random")
```

## Birth-Death Process — A Simple Example

A birth-death process models a population where individuals are born at rate 
$\lambda$ and die at rate $\mu$. Its stoichiometry is:

| Reaction | Reactant | Product | Net change |
|----------|----------|---------|------------|
| Birth    | 1 species | 2 species | +1       |
| Death    | 1 species | 0 species | −1       |

### Define the model

```{r birth-death-model}
# Initial population
x0 <- list(5L)

# Rates: birth=1.0, death=0.5 → net growth rate = 0.5
birth_rate <- 1.0
death_rate <- 0.5

# Rate function (Python-callable via reticulate)
rates <- function(x, t) {
  list(birth_rate * x[[1]], death_rate * x[[1]])
}

# Stoichiometric matrices
nu_reactants <- list(list(1L), list(1L))
nu_products  <- list(list(2L), list(0L))

# Net stoichiometry (product − reactant)
nu <- lapply(seq_along(nu_products), function(i) {
  mapply(function(a, b) as.integer(a - b),
         nu_products[[i]], nu_reactants[[i]], SIMPLIFY = FALSE)
})

stoich <- list(
  nu          = nu,
  DoDisc      = list(1L),
  nuReactant  = nu_reactants,
  nuProduct   = nu_products
)

# Simulation options
my_opts <- list(
  EnforceDo          = list(0L),
  dt                 = 0.01,
  SwitchingThreshold = list(10L)
)

t_max <- 6.0
```

### Run a single trajectory

```{r single-sim}
random$seed(42L)
sim <- jsf$jsf(x0, rates, stoich, t_max,
               config = my_opts, method = "operator-splitting")

# Extract time and population vectors
times <- unlist(sim[[2]])
pop   <- unlist(sim[[1]][[1]])

cat("Simulation produced", length(times), "time points\n")
cat("Final population:", tail(pop, 1), "\n")
```

### Plot the trajectory

```{r plot-single, fig.width=8, fig.height=4.5}
df_single <- data.frame(time = times, population = pop)

ggplot(df_single, aes(x = time, y = population)) +
  geom_step(colour = "#2c7bb6", linewidth = 0.6) +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.2, y = 11.5,
           label = "Switching threshold", hjust = 0, size = 3.2) +
  labs(title = "Birth-Death Process — Single JSF Trajectory",
       subtitle = paste0("λ = ", birth_rate, ", μ = ", death_rate,
                         ", X0 = ", x0[[1]]),
       x = "Time", y = "Population size") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Multiple Trajectories and Mean-Field Comparison

The **mean-field (deterministic) solution** for a birth-death process is 
$X(t) = X_0 \, e^{(\lambda - \mu)\,t}$. We run several stochastic 
trajectories and compare their average to this analytical solution.

```{r multi-sim, cache=TRUE}
set.seed(123)
n_reps <- 100

sims <- lapply(seq_len(n_reps), function(i) {
  random$seed(as.integer(i * 7919L))
  s <- jsf$jsf(x0, rates, stoich, t_max,
               config = my_opts, method = "operator-splitting")
  data.frame(
    time = unlist(s[[2]]),
    pop  = unlist(s[[1]][[1]]),
    rep  = i
  )
})

df_all <- do.call(rbind, sims)
cat("Total data points:", nrow(df_all), "\n")
```

### Compute ensemble average on a regular grid

```{r ensemble-avg}
grid_times <- seq(0, t_max, by = 0.05)

# For each replicate, interpolate onto the common grid
avg_pop <- sapply(grid_times, function(tg) {
  vals <- sapply(seq_len(n_reps), function(i) {
    sub <- sims[[i]]
    idx <- max(which(sub$time <= tg))
    sub$pop[idx]
  })
  mean(vals, na.rm = TRUE)
})

# Mean-field solution
mf_pop <- x0[[1]] * exp((birth_rate - death_rate) * grid_times)

df_avg <- data.frame(
  time     = grid_times,
  avg_jsf  = avg_pop,
  mean_field = mf_pop
)
```

### Plot: Trajectories + Mean-Field

```{r plot-multi, fig.width=9, fig.height=5}
ggplot() +
  # Individual trajectories (light blue)
  geom_step(data = df_all, aes(x = time, y = pop, group = rep),
            colour = "#abd9e9", alpha = 0.25, linewidth = 0.3) +
  # Ensemble average
  geom_line(data = df_avg, aes(x = time, y = avg_jsf, colour = "JSF average"),
            linewidth = 1) +
  # Mean-field
  geom_line(data = df_avg, aes(x = time, y = mean_field, colour = "Mean-field"),
            linewidth = 1, linetype = "dashed") +
  scale_colour_manual(values = c("JSF average" = "#d7191c",
                                 "Mean-field"  = "black")) +
  labs(title = "Birth-Death: 100 JSF Trajectories vs Mean-Field Solution",
       x = "Time", y = "Population size", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = c(0.15, 0.85),
        plot.title = element_text(face = "bold"))
```

## Lotka-Volterra Predator-Prey Model

As a richer example, we simulate the classic predator-prey system using JSF.

```{r lotka-volterra}
x0_lv <- list(50L, 10L)

mA <- 2.00   # prey birth rate
mB <- 0.05   # predation rate
mC <- 1.50   # predator death rate

rates_lv <- function(x, t) {
  list(mA * x[[1]], mC * x[[2]], mB * x[[1]] * x[[2]])
}

nu_react_lv  <- list(list(1L, 0L), list(0L, 1L), list(1L, 1L))
nu_prod_lv   <- list(list(2L, 0L), list(0L, 0L), list(0L, 2L))

nu_lv <- lapply(seq_along(nu_prod_lv), function(i) {
  mapply(function(a, b) as.integer(a - b),
         nu_prod_lv[[i]], nu_react_lv[[i]], SIMPLIFY = FALSE)
})

stoich_lv <- list(
  nu         = nu_lv,
  DoDisc     = list(1L, 1L),
  nuReactant = nu_react_lv,
  nuProduct  = nu_prod_lv
)

opts_lv <- list(
  EnforceDo          = list(0L, 0L),
  dt                 = 0.01,
  SwitchingThreshold = list(30L, 30L)
)

random$seed(42L)
sim_lv <- jsf$jsf(x0_lv, rates_lv, stoich_lv, 10.0,
                  config = opts_lv, method = "operator-splitting")

df_lv <- data.frame(
  time     = unlist(sim_lv[[2]]),
  prey     = unlist(sim_lv[[1]][[1]]),
  predator = unlist(sim_lv[[1]][[2]])
)
```

```{r plot-lv, fig.width=9, fig.height=5}
library(tidyr)

df_lv_long <- pivot_longer(df_lv, cols = c(prey, predator),
                           names_to = "species", values_to = "count")

ggplot(df_lv_long, aes(x = time, y = count, colour = species)) +
  geom_step(linewidth = 0.5) +
  geom_hline(yintercept = 30, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.2, y = 33,
           label = "Switching threshold", hjust = 0, size = 3.2) +
  scale_colour_manual(values = c(prey = "#1a9850", predator = "#d73027"),
                      labels = c("Predator", "Prey")) +
  labs(title = "Lotka-Volterra Predator-Prey — JSF Simulation",
       x = "Time", y = "Population size", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = c(0.85, 0.85),
        plot.title = element_text(face = "bold"))
```

## Session Info

```{r session-info}
sessionInfo()
py_config()
```
